How vanishing web-based tools hamper reproducibility

### 2023-12-13 manuscript/letter.Rmd
2013 data

---
title: "How vanishing web-based tools hamper reproducibility"
date: "`r Sys.Date()`"
output:
  rmdformats::robobook:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    fig_width: 5
    fig_height: 5
bibliography: references.bib
csl: plos-computational-biology.csl
---

Alt title: "No publication without preservation: the argument against web-based tools in genomics"

Alt title: "Time to abandon (most) web-based tools in genomics"

[Anusuiya Bora<sup>1,2</sup>](https://orcid.org/0009-0006-2908-1352),
[Mark Ziemann<sup>1,2</sup>*](https://orcid.org/0000-0002-7688-6974)

@@

Web-based tools are an indispensable resource for conducting genomic and bioinformatics analysis, but if they disappear
it can leave reproducibility gaps in research projects.
Unless such tools allow for preservation of key materials for reproduction, researchers may unwittingly be in breach of
data management mandates.

@@

According to a 2016 survey, 52% of researchers believe we are in a reproducibility crisis [@Baker2016-hm].
In genomics we don't have a lot of data on study reproducibility, but a small survey from 2009 indicates that most
studies are irreproducible due to a lack of shared data, software and documentation [@Ioannidis2009-hv].
The types of tools we use for the analysis also play a role.
Web-based point-and-click tools have become very popular in genomics for the analysis of gene lists, to
interpret whether certain biochemical or signaling pathways are over-represented in a particular genomic study.
These tools are so popular because they do not require the installation of any software, and a gene list analysis
can be conducted in a matter of seconds.
They also require little training to use, which on one hand simplifies bioinformatics tasks, but on the other hand
these tools can easily be misused and results misreported, which can lead to misleading conclusions
[ @Timmons2015-ex ; @Wijesooriya2022-li].
Last year we sought to test whether a small group of articles published in 2019 involving enrichment analysis with the DAVID tool [@Sherman2022-hf]
were reproducible using the authors' own methods.
We found only 4/20 enrichment studies had a high degree of reproduction, while 7/20 had severe discrepancies.
After our pilot study completed we were surprised to hear that the version of DAVID used for all of these studies (v6.8) would no longer be available from June 2022 onwards.
This was disturbing, as according to analysis of PubMed citations, some 20,000 articles citing DAVID enrichment tools will no longer be reproducible with
the original tools, some articles just one or two years old.
This is a prime example of "link decay", a phenomenon where internet based resources are lost over time, which has been raised as a significant and ongoing problem for bioinformatic reproducibility [@Hennessey2013-tj].
Of course DAVID isn't the only webserver without the option of preservation; our analysis of the latest NAR Webserver issue shows 28% (23/81) of webserver tools lacked source code or other features to allow preservation [@Seelow2023-xk].

>>>>

---
title: "How vanishing web-based tools hamper reproducibility"
date: "`r Sys.Date()`"
output:
  rmdformats::robobook:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    fig_width: 5
    fig_height: 5
bibliography: references.bib
csl: genome-biology.csl
---

[Anusuiya Bora<sup>1,2</sup>](https://orcid.org/0009-0006-2908-1352),
[Nhi Nguyen<sup>2</sup>](https://orcid.org/XXXX-XXXX-XXXX-XXXX),
[Mark Ziemann<sup>1,2</sup>*](https://orcid.org/0000-0002-7688-6974)

@@

Web-based tools are an indispensable resource for conducting genomic and bioinformatics analysis, but if they disappear
it can leave reproducibility gaps in research projects.
Unless such tools allow for preservation of key materials for reproduction, researchers may unwittingly be in breach of
data retention mandates.

@@

According to a 2016 survey, 52% of researchers believe we are in a reproducibility crisis [@Baker2016-hm].
In genomics we don't have a lot of data on study reproducibility, but a small survey from 2009 indicates that most
studies are irreproducible due to a lack of shared data, software and documentation [@Ioannidis2009-hv].
The types of tools we use for the analysis also play a role.
Web-based point-and-click tools have become very popular in genomics for the analysis of gene lists, to
interpret whether certain biochemical or signaling pathways are over-represented in a particular genomic study.
These tools are so popular because they do not require the installation of any software, and a gene list analysis
can be conducted in a matter of seconds.
They also require little training to use, which on one hand simplifies bioinformatics tasks, but on the other hand
these tools can easily be misused and results misreported, which can lead to misleading conclusions
[ @Timmons2015-ex ; @Wijesooriya2022-li].
Last year we sought to test whether a small group of articles published in 2019 involving enrichment analysis with the DAVID tool [@Sherman2022-hf]
were reproducible using the authors' own methods.
We found only 4/20 enrichment studies had a high degree of reproduction, while 7/20 had severe discrepancies.
After our pilot study completed we were surprised to hear that the version of DAVID used for all of these studies (v6.8) would no longer be available from June 2022 onwards.
This was disturbing, as according to analysis of PubMed citations, some 20,000 articles citing DAVID enrichment tools will no longer be reproducible with
the original tools, some articles just one or two years old.
This is a prime example of "link decay", a phenomenon where internet based resources are lost over time, which has been raised as a significant and ongoing problem for bioinformatic reproducibility [@Hennessey2013-tj].
Of course DAVID isn't the only webserver without the option of preservation; our analysis of the latest NAR Webserver issue shows 28% (23/81) of webserver tools lacked source code or other features to allow preservation [@Seelow2023-xk].
A look at NAR Webservers from 2013 shows 68% of tools did not provide any preservation options, and 35% (33/95) are no longer online.
